Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
better macro gestures section

A useful way of modelling user interfaces is to consider their
configurations as state vectors and to deploy the methods of matrix
algebra to describe transitions between these states during an
interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions with an interface,
manually coding interaction states  This method has been applied to
usability analysis of resource planning software for security
applications~\cite{Kannampallil:2007fp} where the transition
probability matrix of these state sequences was used to draw
conclusions about the nature of this interactions. In the creative
arts, editing commands at a computer interface for the ``live coding''
of computer music has also been coded as states and analysed using
transition matrices~\cite{Swift:2014tya}, in both the textual and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, rehearsals and studies, we
have used the continuous touch-screen gestures identified by Martin et
al~\cite{Martin:2014cr} as a basis for a musical state-space.
Each performer in each musical piece can then be thought of as
producing a ``score'' of state transitions and an overall transition
matrix for the entire ensemble can be derived by averaging the
individual transition matrices.

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
which can be thought of as Markov process. If there is an (automated)
classification process for extracting each participant's state
sequence from (or during) their interaction data, then we can estimate
the $m \times m$ transition matrix $P$ which characterises this
process by simply counting the number of times state $j$ follows state
$i$. The entries of $P$ are then given by:
$p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}$.
% \begin{equation}
%   p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
% \end{equation}
The transition activity of the whole ensemble can be calculated by
averaging the TM for each performer. The matrix can be normalised so
that the sum of all elements is equal to one.

Previous work~\cite{Martin:2015jk} showed that the raw touch data
obtained from interaction with simple iPad interfaces similar to ours
could be segmented using a Random Forest\cite{Breiman:2001kx}
classifier\footnote{The classifier used was from Python's Scikit-learn
  package\cite{scikit-learn}.}. The classifier was applied to 5-second
windows of the touch-data to provide a chain of gesture-states for
each participant in each session. These chains can be used to create
$9 \times 9$ (the classifier recognises 9~distinct gestures)
transition matrices for individual performers and the whole ensemble
in each session.

>>>>

A useful way of modelling user interfaces is to consider their
theoretical configurations as state vectors and to deploy the methods
of matrix algebra to describe transitions between these states during
an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions and manually code
interaction states to produce a transition model. This method has been
applied to usability analysis of resource planning software for
security applications~\cite{Kannampallil:2007fp} where the transition
probability matrix of these state sequences was used to draw
conclusions about the nature of this interactions. In the creative
arts, editing commands at a computer interface to ``live code''
computer music has also been coded as states and analysed using
transition matrices~\cite{Swift:2014tya}, in both the textual and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, rehearsals and studies, we
have used an automated method to classify logs of touch interaction,
at a regular interval, into nine continuous touch-screen gestures, as
described by Martin et al~\cite{Martin:2015jk}. Each performer in each
musical piece can then be thought of as producing a ``score'' of state
transitions and an overall transition matrix for the entire ensemble
can be derived by averaging the individual transition matrices.

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
process.

If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix $P$ which characterises
this process by setting each entry $p_{ij}$ to be the number of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

We applied a Random Forest~\cite{Breiman:2001kx}
classifier\footnote{The Random Forest classifier used was from
  Python's Scikit-learn package\cite{scikit-learn}.} to 5-second
windows of touch-data in our corpus to provide chains of
gesture-states for each participant in each session following previous
work~\cite{Martin:2015jk}. These chains can be used to create
$9 \times 9$ (the classifier recognises 9~distinct gestures)
transition matrices for individual performers and the whole ensemble
in each session.
